{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-08-06 thesis/tm.tex
smore tm revision

Intel's official TSX documentation summarizes its interface and behaviour in prose \cite{intel-tsx-overview}.
Recent work has used proof assistants to formalize some of the execution semantics
of x86 in general \cite{x86-semantics}
and of transactions in particular (at both hardware- and language-level) \cite{relaxed-transactions-pldi}.
However, state-of-the-art advances in model checking algorithms
still state their theorems and proofs in prose \cite{bpor,optimal-dpor,tsopso,satcheck,mcr,quicksand},
so this section's proofs will regrettably do likewise,
leaving the rigor of mechanization to future work.
The reader may at least draw assurance from the existing formal verifications of
the transactional semantics required herein.

@@

For all the reduction retry sets demonstrated in Table~\ref{tab:tm-verifs},
some inefficiencies remain in its strategy.
For example, it is not clear how to prune soundly
when three or more threads must be reorderd around one transactional preemption point,
or when a second pair of partially-independent transactions interleaves while an existing retry set is already active.
Accordingly, I implemented the optimization as conservatively as possible in these cases,
``saturating'' the retry sets to fall back to no pruning
({\tt update\_hax\_abort\_set()} and
{\tt update\_hax\_abandon\_abort\_set()}, respectively).
Similarly, {\tt xbegin} results other than {\tt \_XABORT\_RETRY}
may depend on the execution logic
(explicit aborts may be conditional on some change by a conflicting thread,
and conflict aborts cannot occur before their conflicting transaction to begin with%
\footnote{See {\tt 410user/progs/htm\_causality.c}; that was a fun realization to have
already halfway into implementing retry sets the wrong way at first.}%
),
so how to soundly prune either success paths or retry aborts
while other abort codes are in play
remains an open problem.
For now, retry sets, true to their name,
%are supported only under {\tt landslide -X},
are incompatible with both the {\tt -A} and {\tt -S} options.

Even so, I must also leave a soundness proof for this optimization to future work.
While
%the basic idea is
motivated by straightforward analogy to the known-sound sleep sets,
its intersection with Landslide's other exploration features may cause unforeseen problems.
For now, its use is prohibited in conjunction with other state-space-affecting features
such as ICB (\sect{\ref{sec:landslide-icb}})
as well as multiple abort codes.
I personally believe retry set reduction to be sound under these restrictions,
having carefully scrutinized its behaviour
while constructing Figure~\ref{fig:retry-set-example}
and from inspecting larger test parameters' state spaces while implementing it;
nevertheless, this falls well short of formal proof.

@@

Section~\ref{sec:tm-design}'s formalization of thread interleavings does not account for read/write reorderings
possible on relaxed consistency architectures \cite{memory-consistency-models}.
In fact,
even after \cite{tm-benchmark-cmu}'s proposed fix to the atomicity protocol in Figure~\ref{fig:htm-fixed},
it is still incorrect on Total Store Order (TSO) architectures such as x86,
let alone on weaker memory models.
Despite stores being totally-ordered, x86 may still reorder stores after subsequent loads
\cite{sully-thesis}.
Accordingly, an execution of lines $8,9a,9b,9c$
may be locally visible to another thread as $9a,8,9b,9c$,
and hence an apparent interleaving of
\[
        \tiat{1},\tjat{1-5},\tiat{7},\underline{\tiat{9a}},\tkat{1-5},\underline{\tiat{8}},\tiat{9b-B}
\]
is possible
(reordered accesses underlined for emphasis).
An acquire barrier is needed between lines 8 and 9 to solve this problem on TSO \cite{tsx-need-barrier}
(on x86, either {\tt mfence} or {\tt xchg}/{\tt xadd}).
% gcc generates the latter for its __sync_lock_test_and_set intrinsic
Recent work \cite{relaxed-transactions-pldi} also demonstrated unsoundness
in a similar lock elision implementation on ARMv8 (PSO),
in which the transactional path reads the lock's internal state directly
rather than using a separate flag.
in Figure~\ref{fig:htm-fixed}, a release barrier before line A is also necessary under PSO.
% On Partial Store Order (PSO) architectures,
% %even more barriers may be necessary.
% a release barrier is necessary before line A as well.
% Recent work \cite{relaxed-transactions-pldi} also discovered similar unsoundness
% in lock elision \cite{speculative-lock-elision}
% on ARMv8 (PSO).
% Recent work \cite{relaxed-transactions-pldi} also discovered similar unsoundness

>>>>

Intel's official TSX documentation summarizes its interface and behaviour in prose \cite{intel-tsx-overview}.
Recent work has used proof assistants to formalize some of the execution semantics
of x86 in general \cite{x86-semantics}
and of transactions in particular (at both hardware- and language-level) \cite{relaxed-transactions-pldi}.
However, state-of-the-art advances in model checking algorithms
still state their theorems and proofs in prose \cite{bpor,optimal-dpor,tsopso,satcheck,mcr,quicksand,cpp-model-checking},
so this section's proofs will regrettably do likewise,
leaving the rigor of mechanization to future work.
The reader may at least rest assured that the proofs herein
rely on transactional semantics that have themselves been formally verified.

@@

For all the reduction retry sets demonstrated in Table~\ref{tab:tm-verifs},
some inefficiencies remain in its strategy.
For example, it is not clear how to prune soundly
when three or more threads must be reorderd around one transactional preemption point,
or when a second pair of partially-independent transactions interleaves while an existing retry set is already active.
Accordingly, I implemented the optimization as conservatively as possible in these cases,
``saturating'' the retry sets to fall back to no pruning
({\tt update\_hax\_abort\_set()} and
{\tt update\_hax\_abandon\_abort\_set()}, respectively).

Likewise, the cases of {\tt htm2}(2,3) and (4,1),
in which retry sets achieved better reduction than STM mode,
show that the latter does not necessarily subsume the former,
and that combining the two could in theory achieve further reduction still.
However, {\tt xbegin} results other than {\tt \_XABORT\_RETRY}
may depend on the execution logic
(explicit aborts may be conditional on some change by a conflicting thread,
and conflict aborts cannot occur before their conflicting transaction to begin with%
\footnote{See {\tt 410user/progs/htm\_causality.c}; that was a fun realization to have
already halfway into implementing retry sets the wrong way at first.}%
),
so how to soundly prune either success paths or retry aborts
while other abort codes are in play
remains an open problem.

While
%the basic idea is
motivated by straightforward analogy to the known-sound sleep sets,
the intersection of retry sets with Landslide's other exploration features may cause unforeseen problems.
For now, its use is prohibited in conjunction with other state-space-affecting features
such as ICB (\sect{\ref{sec:landslide-icb}})
as well as multiple abort codes.
I personally believe retry set reduction to be sound under these restrictions,
having carefully scrutinized its behaviour
while constructing Figure~\ref{fig:retry-set-example}
and from inspecting state spaces arising from larger test parameters as well; %while implementing it;
nevertheless, this falls well short of formal proof, which I must leave to future work.

@@

Section~\ref{sec:tm-design}'s formalization of thread interleavings does not account for read/write reorderings
possible on relaxed consistency architectures \cite{memory-consistency-models}.
In fact,
even after \cite{tm-benchmark-cmu}'s proposed fix to the atomicity protocol in Figure~\ref{fig:htm-fixed},
it is still incorrect on Total Store Order (TSO) architectures such as x86,
let alone on weaker memory models.
Despite stores being totally-ordered, x86 may still reorder stores after subsequent loads
\cite{sully-thesis}.
Accordingly, an execution of lines $8,9a,9b,9c$
may be locally visible to another thread as $9a,8,9b,9c$,
and hence an apparent interleaving of
\[
        \tiat{1},\tjat{1-5},\tiat{7},\underline{\tiat{9a}},\tkat{1-5},\underline{\tiat{8}},\tiat{9b-B}
\]
is possible
(reordered accesses underlined for emphasis).
An acquire barrier is needed between lines 8 and 9 to solve this problem on TSO \cite{tsx-need-barrier}
(on x86, either {\tt mfence} or {\tt xchg}/{\tt xadd}).
% gcc generates the latter for its __sync_lock_test_and_set intrinsic
Recent work \cite{relaxed-transactions-pldi} also demonstrated unsoundness
in a similar lock elision implementation on ARMv8 (PSO),
in which the transactional path reads the lock's internal state directly
rather than using a separate flag.
In Figure~\ref{fig:htm-fixed}, a release barrier before line A is also necessary under PSO.
% On Partial Store Order (PSO) architectures,
% %even more barriers may be necessary.
% a release barrier is necessary before line A as well.
% Recent work \cite{relaxed-transactions-pldi} also discovered similar unsoundness
% in lock elision \cite{speculative-lock-elision}
% on ARMv8 (PSO).
% Recent work \cite{relaxed-transactions-pldi} also discovered similar unsoundness
